Testing the motor and cognitive foundations of Paleolithic social transmission

### 2021-08-16 manuscript/manuscript.Rmd
all figures are updated and inserted properly in the main text.

Studies investigating these questions have used a range of different experimental designs (e.g., varying technological goals/instructions, training times, raw materials, live vs. recorded instruction, lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached disparate conclusions regarding the neurocognitive and social foundations of skill acquisition. It is plausible that these discordant results reflect actual diversity in how humans acquire and master stone tool making skills. However, this failure of results to generalize across artificial experimental manipulations [[cf](cf.%20Yarkoni%202020). @yarkoni2020] also raises doubts regarding the external validity [@eren2016] of conclusions with respect to real-world Paleolithic learning contexts. To address this, we conducted an exploratory study that draws on lessons from previous research in an attempt to balance the pragmatic and theoretical tradeoffs inherent in experimental studies of stone knapping skill acquisition [@pargeter2019; @stout2015].

@@

After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour. Participants were instructed not to seek out additional training or information on stone tool making (i.e., via the internet) during these breaks. Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching. In both conditions, participants were first given an opportunity to inspect and handle examples (**Fig. \@ref(fig:fig1)**) of the kind of stone tools (flakes) they are being asked to produce. They were told that their objective was to produce as many flakes as possible from the materials provided. This meant that even the untaught condition included some minimal instruction (being told the objective) , however this was considered to be unavoidable without creating a much more elaborate and naturalistic context in which participants would develop their own technological goals. Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.
```{r fig1, echo=FALSE, out.width = '75%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment. The demonstration flakes were made from the same basalt as used in the experiment with the same knapping technique."}
knitr::include_graphics(here::here("figure/Fig1.PNG"))
```

@@

Each participant was provided with 9 cores for use over the 2-hour experiment. These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer. This produced irregular, angular chunks for use in the experiment, weighing between 459g - 1876g (mean = 975g). All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production. Cores were sorted by shape and weight and then distributed evenly to each participant. As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided. This was also true comparing the two experimental conditions (Taught vs. Untaught mean weight = 1001g vs. 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs. 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p = 0.05). Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.

@@

In both conditions, three researchers were present to record activities and collect materials. Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras (Figure 1). Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred. However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later. The order of cores used and associated hammerstones were recorded for each participant during the experiment.

In the untaught condition, a researcher (DS) sat with the participants and made stone tools but remained silent and made no effort to facilitate learning (e.g., through gesture, modified performance, facial expression, attention direction, or verbal instruction). Over the 2-hour period, the researcher completely reduced four cores (one every \~30 minutes). Participants were not restricted from talking to each other, as this would create an unnatural and potentially stressful social context that might affect learning. Participants were asked to avoid any form of communication about the tool making task specifically, and they complied with this request. Participants in this condition thus had the opportunity to observe tool making by an expert and/or by other learners, should they choose to do so, but received no intentional instruction.

@@

This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (Figures). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (Fig. Xa), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj. $R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj. $R^2$ = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj. $R^2$ = 0.7), but no relation in the Trained group (Fig. Xb). A qualitatively similar trend with respect to Quality vs. Quantity (Fig. Xc) did not achieve significance (F[1, 17] = 0.6, p = 0.2).

@@

@@

Table summarizes the results of ANOVA tests group level difference in central tendency on various performance measures. We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (Figure). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (Figures-Core examples too). We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.

While we did not find significant differences in central tendency between our two experimental groups, results (Figure and Figure) did indicate lower variance in the trained group. To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances. Table and Figure present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass). In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times. The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (Figure), rather than to shift the mean.

@@

Table shows no significant training effects across the two performance measures either as grouped data or between individuals (Figures). This result demonstrated that flaking outcomes did not change dramatically across the study interval. This lack of significant learning effects is confirmed by an inspection of individual learning curves (Figure). The one significant main training effect related to core starting mass (with a strong effect size = 0.25). On average, core starting masses start low and increase, showing that participants selected smaller nodules first. As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules. This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure).

@@

### Model 1: Individual differences and quantity flaking

@@

Table presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (Figure), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.

Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (Figures). In each case, higher scores were associated with better performance in the uninstructed group but worse performance in the instructed group. Positive effects in the uninstructed group were as expected, given the hypothesized importance of spatial cognition (Coolidge and Wynn 2005) and social learning (Morgan et al. 2015) in the acquisition of knapping skills. Negative effects in the trained group are unexpected but presumably reflect differences in learning strategies adopted under instruction.

@@

The Quantity model roughly paralleled results for Total Production, yielding substantial and significant interactions between training condition, n-back level, BEAST scores, and grip strength. As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the trained group but higher or unchanged Quantity in the untrained group (Figure).

Unlike Total Productivity, the effect of Grip Strength on per-core Quantity was mediated by an interaction with learning condition (Figure). Thus, high Grip Strength enabled individuals in both groups to produce more total debitage, but only Instructed individuals translated Grip Strength into more intense reduction of individual cores, including not only delta mass, but also number and proportion of larger pieces.

@@

The Quality model did produce two statistically (p\< 0.05) significant interaction effects (RPM \* BEAST & Fitts \* n-back). However, these interactions had relatively small effects on Quality (\<0.5) and we believe that interpreting these results from our small, exploratory study would be inappropriate. Figure shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.

@@

We designed this exploratory study primarily to trial experimental design elements such as training time, conditions, and raw materials and to collect preliminary data on the effect of individual differences and training on knapping outcomes. We thus focused on collecting quantitative psychometric and lithic data. However, we also considered that quantifying participant knapping behaviors as well as products could be important for future studies. To support methods development in this regard, we made ad hoc notes on observed behaviors during the experiments and video-recorded all experiments to enable later, more systematic analyses yet to be completed. However, even casual behavior observation was sufficient to reveal an unexpected effect. Whereas all trained participants copied the general posture and technique of the expert (free hand knapping seated in a chair) fully half (6) of the uninstructed participants experimented with or even knapped all of their cores using the floor as a support (Figure). Three of these participants were in the same session, which is also the only group composed of just three individuals. In this group, knapping on the ground appears to have been transmitted from one participant to the other two based on appearance order and the point of gaze of participants.

@@

We also found a strong positive effect of Grip Strength on Total Productivity independent of learning condition (Figure). While it is tempting to interpret this with respect to the demands for hand strength specifically, it is important to remember that grip strength is strongly correlated with total muscle strength [@wind2010] and overall fitness [@sasaki2007]. Thus, it is best taken to indicate some importance of fitness generally in increasing the rate and intensity of core reduction by naïve knappers, potentially affecting rate of work and the kinetic energy of the swing as well as the handling of core and hammerstone. It thus provides further support for hypotheses positing stone tool making as a selection pressure on the functional anatomy of hand, arm, and shoulder [e.g., @williams-hatala2018], but initially appears orthogonal to variations in learning condition and knapping behaviors in our study. However, we also found that the effect of Grip Strength on per-core knapping Quantity is dependent on teaching (Figure). The absence of this effect in the uninstructed group reflects the weaker association between Total Productivity and per-core Quality across these participants (Fig.Xa) and shows Grip Strength increased uninstructed Productivity specifically by allowing them to knap more cores rather than to reduce individual cores more heavily. In keeping with this, uninstructed Grip Strength is positively correlated with Total Cores knapped (r2 = 0.54, p = 0.01). Conversely, strength allowed instructed participants to increase their average Quantity per core without affecting the total number of cores knapped (r2 = 0.18, p = 0.165). Thus, strength appears to have achieved its effects on core reduction rate and intensity in different ways, depending on teaching. This difference is likely related to the homogenizing effect of teaching on knapping rate (all instructed participants knapped 6 or more cores) and methods. Subjectively, knapping behaviors of uninstructed participants often appeared more physically demanding (e.g., greater number of non-productive blows, rapid and unregulated battering) which would imply different demands on both strength and aerobic fitness [@williams-hatala2021; @mateos2019]. However, this remains to be systematically investigated.

>>>>

Studies investigating these questions have used a range of different experimental designs (e.g., varying technological goals/instructions, training times, raw materials, live vs. recorded instruction, lithic/skill assessment metrics, pseudo-knapping tasks etc.) and reached disparate conclusions regarding the neurocognitive and social foundations of skill acquisition. It is plausible that these discordant results reflect actual diversity in how humans acquire and master stone tool making skills. However, this failure of results to generalize across artificial experimental manipulations [cf. @yarkoni2020] also raises doubts regarding the external validity [@eren2016] of conclusions with respect to real-world Paleolithic learning contexts. To address this, we conducted an exploratory study that draws on lessons from previous research in an attempt to balance the pragmatic and theoretical tradeoffs inherent in experimental studies of stone knapping skill acquisition [@pargeter2019; @stout2015].

@@

After individual difference testing, participants engaged in a 2-hour stone tool making session, with a 15-minute break after 1 hour. Participants were instructed not to seek out additional training or information on stone tool making (i.e., via the internet) during these breaks. Each group of participants was randomly assigned to one of two experimental conditions: no teaching or teaching. In both conditions, participants were first given an opportunity to inspect and handle examples (**Figure** \@ref(fig:fig1)) of the kind of stone tools (flakes) they are being asked to produce. They were told that their objective was to produce as many flakes as possible from the materials provided. This meant that even the untaught condition included some minimal instruction (being told the objective) , however this was considered to be unavoidable without creating a much more elaborate and naturalistic context in which participants would develop their own technological goals. Such a design would also be expected to increase behavioral variability, demanding correspondingly larger samples of participants to identify patterns and making direct comparisons with the taught condition.

```{r fig1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment. The demonstration flakes were made from the same basalt as used in the experiment with the same knapping technique."}
knitr::include_graphics(here::here("figure/Figure1.jpg"))
```

@@

Each participant was provided with 9 cores for use over the 2-hour experiment. These cores were produced from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer. This basalt has not been mechanically compared [e.g., rebound hardness, @braun2009] to East African basalts, but appears qualitatively similar to finer grained examples from sites around Lake Turkana (D. Braun, pers. comm.) and at Gona. Spalling produced irregular, angular chunks (**Figure** \@ref(fig:fig2)) for use in the experiment, weighing between 459g - 1876g (mean = 975g). All cores were weighed, measured (Length, Width, Thickness), and painted white so that new fracture surfaces could be discriminated from those created during production. Cores were sorted by shape and weight and then distributed evenly to each participant. As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = .6, df1=22, df2 = 184, p = 0.9) of cores provided. This was also true for cores provided to participants across the two experimental conditions (Taught vs. Untaught mean weight = 1001g vs. 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs. 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p= 0.05). Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores actually used across participants and conditions could still emerge as a result of selection bias.

```{r fig2, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="3D scan of a basalt core prior to knapping. Cores were spray painted white to facilitate subsequent technological analyses of the flake scars and flaking intensity. Scale in the background is in 10mm blocks."}
knitr::include_graphics(here::here("figure/Figure2.jpg"))
```

@@

In both conditions, three researchers were present to record activities and collect materials. Participants were seated in a circle [@shea2015] and experiments were video recorded using two cameras. Participants were free to select hammerstones from the common pile and to work on any or all of their nine assigned cores in any order they preferred (**Figure** \@ref(fig:fig3)). However, each core and all associated debitage were collected before participants were allowed to start working on a new core, so it was not possible to partially work and then return to a particular core later. The order of cores used and associated hammerstones were recorded for each participant during the experiment.

```{r fig3, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects selecting hammerstones."}
knitr::include_graphics(here::here("figure/Figure3.jpg"))
```

In the untaught condition, a researcher (DS) sat with the participants and made stone tools but remained silent and made no effort to facilitate learning (e.g., through gesture, modified performance, facial expression, attention direction, or verbal instruction). Over the 2-hour period, the researcher completely reduced four cores (one every \~30 minutes). Participants were not restricted from talking to each other, as this would create an unnatural and potentially stressful social context that might affect learning. Participants were asked to avoid any form of communication about the tool making task specifically, and they complied with this request. Participants in this condition thus had the opportunity to observe tool making (**Figure** \@ref(fig:fig4)) by an expert and/or by other learners, should they choose to do so, but received no intentional instruction.

```{r fig4, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects observing expert knapper in the untaught condition."}
knitr::include_graphics(here::here("figure/Figure4.jpg"))
```

@@

This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (**Figure** \@ref(fig:fig5) and **Figure** \@ref(fig:fig6)). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (**Figure** \@ref(fig:fig5)a), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj. $R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj. $R^2$ = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj. $R^2$ = 0.7), but no relation in the Trained group (**Figure** \@ref(fig:fig5)b). A qualitatively similar trend with respect to Quality vs. Quantity (**Figure** \@ref(fig:fig5)c) did not achieve significance (F[1, 17] = 0.6, p = 0.2).

```{r fig5, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Relationships between flaking quantity, quality, productivity, and the two training conditions. Each dot represents a participant, colors represent training conditions."}
knitr::include_graphics(here::here("figure/Figure5.jpg"))
```

@@

```{r fig6, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparison of untrained and trained detached pieces illustrating the ability of Trained participants to maximize Quality and Productivity at the same time."}
knitr::include_graphics(here::here("figure/Figure6.jpg"))
```

@@

Table summarizes the results of ANOVA tests group level difference in central tendency on various performance measures. We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (**Figure** \@ref(fig:fig7)). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (**Figure** \@ref(fig:fig8)). We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.

```{r fig7, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Results showing significant differences between the instructor (expert) and novice flaking performance."}
knitr::include_graphics(here::here("figure/Figure7.png"))
```

```{r fig8, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparison of untaught, taught, and expert cores."}
knitr::include_graphics(here::here("figure/Figure8.jpg"))
```

While we did not find significant differences in central tendency between our two experimental groups, results (**Figure** \@ref(fig:fig7)) did indicate lower variance in the trained group. To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances. Table and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass). In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times. The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.

```{r fig9, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Variance comparisons between trained and untrained individuals across various flaking performance metrics. Note the lower variance of the trained group in all the plots."}
knitr::include_graphics(here::here("figure/Figure9.jpg"))
```

@@

Table shows no significant training effects across the two performance measures either as grouped data or between individuals (**Figure** \@ref(fig:fig10) and **Figure** \@ref(fig:fig11)). This result demonstrated that flaking outcomes did not change dramatically across the study interval. This lack of significant learning effects is confirmed by an inspection of individual learning curves (Figure). The one significant main training effect related to core starting mass (with a strong effect size = 0.25). On average, core starting masses start low and increase, showing that participants selected smaller nodules first. As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules. This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure???).

```{r fig10, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparisons of nodule starting mass across the study period by training condition. Results show a significant relationship between nodule starting mass and length of time in the study."}
knitr::include_graphics(here::here("figure/Figure10.jpg"))
```

```{r fig11, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Individual learning curves derived from each subject’s relative core use order and the two flake performance factors."}
knitr::include_graphics(here::here("figure/Figure11.jpg"))
```

@@

### Model 1: Individual differences and overall productivity

@@

Table presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.

```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
knitr::include_graphics(here::here("figure/Figure12.jpeg"))
```

Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (**Figure** \@ref(fig:fig13)). In each case, higher scores were associated with better performance in the uninstructed group but worse performance in the instructed group. Positive effects in the uninstructed group were as expected, given the hypothesized importance of spatial cognition (Coolidge and Wynn 2005) and social learning (Morgan et al. 2015) in the acquisition of knapping skills. Negative effects in the trained group are unexpected but presumably reflect differences in learning strategies adopted under instruction.

```{r fig13, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Two significant interaction effects of training condition, social information, and visuo-spatial working memory on overall flaking productivity."}
knitr::include_graphics(here::here("figure/Figure13.png"))
```

@@

The Quantity model roughly paralleled results for Total Production, yielding substantial and significant interactions between training condition, n-back level, BEAST scores, and grip strength. As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the trained group but higher or unchanged Quantity in the untrained group (**Figure** \@ref(fig:fig14)).

```{r fig14, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Two significant interaction effects of training condition, visuo-spatial working memory, and social information, on overall flaking quality."}
knitr::include_graphics(here::here("figure/Figure14.jpg"))
```

Unlike Total Productivity, the effect of Grip Strength on per-core Quantity was mediated by an interaction with learning condition (**Figure** \@ref(fig:fig15)). Thus, high Grip Strength enabled individuals in both groups to produce more total debitage, but only Instructed individuals translated Grip Strength into more intense reduction of individual cores, including not only delta mass, but also number and proportion of larger pieces.

```{r fig15, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Three significant interaction effects of training condition, visuo-spatial working memory and social information, and training and grip strength use on quantity flaking."}
knitr::include_graphics(here::here("figure/Figure15.jpg"))
```

@@

The Quality model did produce two statistically (p\< 0.05) significant interaction effects (RPM \* BEAST & Fitts \* n-back). However, these interactions had relatively small effects on Quality (\<0.5) and we believe that interpreting these results from our small, exploratory study would be inappropriate. **Figure** \@ref(fig:fig16) shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.

```{r fig16, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant interactions with low estimated effect sizes from the quality flaking model."}
knitr::include_graphics(here::here("figure/Figure16.jpg"))
```

@@

We designed this exploratory study primarily to trial experimental design elements such as training time, conditions, and raw materials and to collect preliminary data on the effect of individual differences and training on knapping outcomes. We thus focused on collecting quantitative psychometric and lithic data. However, we also considered that quantifying participant knapping behaviors as well as products could be important for future studies. To support methods development in this regard, we made ad hoc notes on observed behaviors during the experiments and video-recorded all experiments to enable later, more systematic analyses yet to be completed. However, even casual behavior observation was sufficient to reveal an unexpected effect. Whereas all trained participants copied the general posture and technique of the expert (free hand knapping seated in a chair) fully half (6) of the uninstructed participants experimented with or even knapped all of their cores using the floor as a support (**Figure** \@ref(fig:fig17)). Three of these participants were in the same session, which is also the only group composed of just three individuals. In this group, knapping on the ground appears to have been transmitted from one participant to the other two based on appearance order and the point of gaze of participants.

```{r fig17, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Novices in the untaught condition developing and transmitting a form of bipolar technology involving core reduction on the floor."}
knitr::include_graphics(here::here("figure/Figure17.jpg"))
```

@@

We also found a strong positive effect of Grip Strength on Total Productivity independent of learning condition (Figure). While it is tempting to interpret this with respect to the demands for hand strength specifically, it is important to remember that grip strength is strongly correlated with total muscle strength [@wind2010] and overall fitness [@sasaki2007]. Thus, it is best taken to indicate some importance of fitness generally in increasing the rate and intensity of core reduction by naïve knappers, potentially affecting rate of work and the kinetic energy of the swing as well as the handling of core and hammerstone. It thus provides further support for hypotheses positing stone tool making as a selection pressure on the functional anatomy of hand, arm, and shoulder [e.g., @williams-hatala2018], but initially appears orthogonal to variations in learning condition and knapping behaviors in our study. However, we also found that the effect of Grip Strength on per-core knapping Quantity is dependent on teaching (Figure). The absence of this effect in the uninstructed group reflects the weaker association between Total Productivity and per-core Quality across these participants (Fig.Xa) and shows Grip Strength increased uninstructed Productivity specifically by allowing them to knap more cores rather than to reduce individual cores more heavily. In keeping with this, uninstructed Grip Strength is positively correlated with Total Cores knapped ($R^2$ = 0.54, p = 0.01). Conversely, strength allowed instructed participants to increase their average Quantity per core without affecting the total number of cores knapped ($R^2$ = 0.18, p = 0.165). Thus, strength appears to have achieved its effects on core reduction rate and intensity in different ways, depending on teaching. This difference is likely related to the homogenizing effect of teaching on knapping rate (all instructed participants knapped 6 or more cores) and methods. Subjectively, knapping behaviors of uninstructed participants often appeared more physically demanding (e.g., greater number of non-productive blows, rapid and unregulated battering) which would imply different demands on both strength and aerobic fitness [@williams-hatala2021; @mateos2019]. However, this remains to be systematically investigated.
